---
title: "Power analysis"
author: "Mikhail Dozmorov"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: united
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is') #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r}
# http://gastonsanchez.com/how-to/2015/05/15/normal-curve/
# normal function 
normal <- function(x, mu = 0, sigma = 1) {
  # x: values on x-axis
  # mu: mean value
  # sigma: standard deviation
  
  # output
  1 / (sigma * sqrt(2*pi)) * exp(-0.5 * ((x - mu)/ sigma)^2)
}
```


# Example

You are working on designing a new cancer drug and want to increase its half life in the body from currend standard of 20 hours (with a std. dev. of 4 hours) for cancer drugs.

You believe your new design can increase the half-life by 2 hours to 22 hours. You have 44 random samples of your new drug for half life.

Determine the power of your study (for $\alpha = 0.05$). If it is lacking power, what can you do to increase its power?

## Components of power analysis

- Model (test)
- Test Effect (effect size and variability)
- Sample size (n)
- Test size (significance level, $\alpha$)
- Power of test ($1-\beta$)

## Hypotheses

- $H_0: \; \mu = 20$
- $H_A: \; \mu = 22$

**Effect size** 22-20=2

Two-tailed t-test

## At what point on $H_0$ distribution does rejection region start (corresponds to $\alpha/2$)?

T-table for df=43. $t_{critical}value=$ `r abs(qt(0.025, 43))`

$$t_{stat} = \frac{\hat{x} - \mu}{\frac{s}{\sqrt{n}}}$$

`r abs(qt(0.025, 43))` $= \frac{x_{critical} - 20}{4/\sqrt{44}}$

$x_{critical}$ = 21.2161

## $H_0$ distribution

Area to the right corresponds to $\alpha / 2$

```{r fig.height=3}
# x-y coordinates for Standard normal curve
x <- seq(15, 25, length.out = 500)
y <- normal(x, mu = 20, sigma = 1)

# standard normal curve plot
op = par(mar = c(2.5,1,1,1))
plot(x, y, type = 'l', col = '#00ABFF', lwd = 3,
     axes = FALSE, xlab = '', ylab = '')
axis(side = 1, at = seq(15, 25, by = 1))
abline( v = 21.2161)
```

## $H_A$ distribution

Area to the right corresponds to $AUC = \beta$ - probability of a Type II error. 

To the right - $1 - \beta$ - power

```{r fig.height=3}
# x-y coordinates for Standard normal curve
x <- seq(15, 25, length.out = 500)
y <- normal(x, mu = 22, sigma = 1)

# standard normal curve plot
op = par(mar = c(2.5,1,1,1))
plot(x, y, type = 'l', col = '#00ABFF', lwd = 3,
     axes = FALSE, xlab = '', ylab = '')
axis(side = 1, at = seq(15, 25, by = 1))
abline( v = 21.2161)
```

## How to find $\beta$?

What is the probability of having a value less than $t_{critical}$ (21.2161) in $H_A$ distribution?

$t_{\beta} = (21.2161 - 22) / (4/\sqrt{44}) = -1.299$

Plot this value on a t-distribution

```{r fig.height=3}
# x-y coordinates for Standard normal curve
x <- seq(-5, 5, length.out = 500)
y <- normal(x, mu = 0, sigma = 1)

# standard normal curve plot
op = par(mar = c(2.5,1,1,1))
plot(x, y, type = 'l', col = '#00ABFF', lwd = 3,
     axes = FALSE, xlab = '', ylab = '')
axis(side = 1, at = seq(-5, 5, by = 1))
abline( v = -1.299)
```

$\beta = P(t < -1.299) = P(t>1.299) =$ `r pnorm(-1.299, 0, 1)` (0.1003 in the video)

$Power = 1 - \beta = 1 - 0.1003 = 0.8997$

Power > 0.8, therefore experimental design is powerful. Should be able to detect effect size.

## References

See power analysis in action: "Power analysis example" video [https://www.youtube.com/watch?v=fRm2dEWSJrk](https://www.youtube.com/watch?v=fRm2dEWSJrk) by Matthew Novak.

- https://www.bu.edu/orccommittees/iacuc/policies-and-guidelines/sample-size-calculations/

- http://www.3rs-reduction.co.uk/html/6__power_and_sample_size.html

- http://www.ats.ucla.edu/stat/seminars/Intro_power/

- http://biostat.mc.vanderbilt.edu/wiki/Main/PowerSampleSize

### Online tools

- http://www.gpower.hhu.de/en.html

- http://powerandsamplesize.com/

- http://www.sample-size.net/sample-size-survival-analysis/

- https://www.stat.ubc.ca/~rollin/stats/ssize/n2.html

- http://scotty.genetics.utah.edu/

### R packages

- `ssize.fdr` - Sample Size Calculations for Microarray Experiments, https://cran.r-project.org/web/packages/ssize.fdr/index.html

- `ssize` - Estimate Microarray Sample Size, https://bioconductor.org/packages/release/bioc/html/ssize.html

- `sizepower` - Sample Size and Power Calculation in Micorarray Studies, https://bioconductor.org/packages/release/bioc/html/sizepower.html

- `OCplus` - Operating characteristics plus sample size and local fdr for microarray experiments, https://bioconductor.org/packages/release/bioc/html/OCplus.html

- `FDRsampsize` - Compute Sample Size that Meets Requirements for Average Power and FDR, https://cran.r-project.org/web/packages/FDRsampsize/index.html

- `SSPA` - General Sample Size and Power Analysis for Microarray and Next-Generation Sequencing Data, https://bioconductor.org/packages/release/bioc/html/SSPA.html

- `RNASeqPower` - Sample size for RNAseq studies, https://bioconductor.org/packages/release/bioc/html/RNASeqPower.html