---
title: "MDS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Source](https://personality-project.org/r/mds.html)

Given a set of distances (dis-similarities) between objects, is it possible to recreate a dimensional representation of those objects?

**Model**: Distance = square root of sum of squared distances on k dimensions dxy = ÷(xi-yi)2

**Data**: a matrix of distances

Find the dimensional values in k = 1, 2, ... dimensions for the objects that best reproduces the original data.

**Example**: Consider the distances between nine American cities. Can we represent these cities in a two dimensional space.

```{r}
library(psych)
data(cities)
cities
```

The output gives us the the original distance matrix (just to make sure we put it in correctly, the x,y coordinates for each city, and then the following graph.

```{r}
city.location <- cmdscale(cities, k=2)    #ask for a 2 dimensional solution
round(city.location,0)        #print the locations to the screen
```

This solution can be represented graphically:

```{r}
plot(city.location,type="n", xlab="Dimension 1", ylab="Dimension 2",main ="cmdscale(cities)")    #put up a graphics window
text(city.location,labels=names(cities))     #put the cities into the map
```

Note that the solution is not quite what we expected (it is giving us a mirrored Australian orientation to American cities.) However, by reversing the signs in city.location, we get the more conventional representation:

```{r}
city.location <- -city.location
plot(city.location,type="n", xlab="Dimension 1", ylab="Dimension 2",main ="cmdscale(cities)")    #put up a graphics window
text(city.location,labels=names(cities))     #put the cities into the map
```

A useful feature is R is most commands have an extensive help file. Asking for help(cmdscale) shows that R includes a distance matrix for 20 European cities. The following commands (taken from the help file) produce a nice two dimensional solution. (Note that since dimensions are arbitrary, the second dimension needs to be flipped to produce the conventional map of Europe.)

```{r}
loc <- cmdscale(eurodist, k = 2)
x <- loc[,1]
y <- -loc[,2]
plot(x, y, type="n", xlab="", ylab="", main="cmdscale(eurodist)")
text(x, y, colnames(as.matrix(eurodist)), cex=0.8)
```

For gene expression matrices, use `limma::plotMDS` function, [http://web.mit.edu/~r/current/arch/i386_linux26/lib/R/library/limma/html/plotMDS.html](http://web.mit.edu/~r/current/arch/i386_linux26/lib/R/library/limma/html/plotMDS.html)

