---
title: "Alignment algorithms"
# linktitle: "1. Introduction to RStudio and Images"
date: "2021-02-17"
class_date: "2021-02-17"
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 6
type: docs
# weight: 1 
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Slides

### Alignment introduction

<a href="../../slides/04_Alignment/01_Alignment_intro.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/04_Alignment/01_Alignment_intro.html#1')
```

### Alignment introduction

<a href="../../slides/04_Alignment/01_Alignment_intro.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/04_Alignment/01_Alignment_intro.html#1')
```

### Approximate matching

<a href="../../slides/04_Alignment/02_Approximate_matching.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/04_Alignment/02_Approximate_matching.html#1')
```

[02_Approximate_matching_Langmead.pdf](../../slides/04_Alignment/02_Approximate_matching_Langmead.pdf)

- "[Approximate matching, Hamming and edit distance](https://www.youtube.com/watch?v=MCvHeW13DsE&index=30&list=PL2mpR0RYFQsBiCWVJSvVAO3OJ2t7DzoHA)"
    - [naive_Hamming.R](../../slides/04_Alignment/lab/naive_Hamming.R)
- "[Solving the edit distance problem](https://www.youtube.com/watch?v=8Q2IEIY2pDU&index=33&list=PL2mpR0RYFQsBiCWVJSvVAO3OJ2t7DzoHA)"
    - [edDistRecursive.R](../../slides/04_Alignment/lab/edDistRecursive.R)
- "[Using dynamic programming for edit distance](https://www.youtube.com/watch?v=0KzWq118UNI&index=34&list=PL2mpR0RYFQsBiCWVJSvVAO3OJ2t7DzoHA)"
    - [edDistDynamic.R](../../slides/04_Alignment/lab/edDistDynamic.R)
- "[Edit distance for approximate matching](https://www.youtube.com/watch?v=NjfNZzJiu8o&list=PL2mpR0RYFQsBiCWVJSvVAO3OJ2t7DzoHA&index=36)"
- "[Meet the family: global and local alignment](https://www.youtube.com/watch?v=-bjSPP2v6_Q&index=37&list=PL2mpR0RYFQsBiCWVJSvVAO3OJ2t7DzoHA)"

### Needleman-Wunsch global alignment

<a href="../../slides/04_Alignment/03_Needleman_Wunsch.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/04_Alignment/03_Needleman_Wunsch.html#1')
```

### Smith-Waterman local alignment

<a href="../../slides/04_Alignment/04_Smith-Waterman.pdf" target="_blank">Lecture slides in PDF format</a>


