---
title: "Bioconductor"
# linktitle: "1. Introduction to RStudio and Images"
date: "2021-02-22"
class_date: "2021-02-22"
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 7
type: docs
# weight: 1 
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Slides

### Bioconductor introduction

<a href="../../slides/05_Bioconductor/01_Bioconductor_introduction.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/01_Bioconductor_introduction.html#1')
```

<a href="http://www.haowulab.org/teaching/bioc/biostrings.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/biostrings.pdf')
```

<a href="../../slides/05_Bioconductor/lab/biostrings.R" target="_blank">biostrings.R</a> - BioStrings demo

### Genomic ranges

<a href="../../slides/05_Bioconductor/02_GRanges.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/02_GRanges.html#1')
```

<a href="https://github.com/haowulab/haowulab.github.io/blob/master/teaching/bioc/genomicRanges.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/genomicRanges.pdf')
```

<a href="../../slides/05_Bioconductor/lab/genomicRanges.R" target="_blank">genomicRanges.R</a> - GenomicRanges demo. Other demos: <a href="../../slides/05_Bioconductor/lab/ranges.Rmd" target="_blank">ranges.Rmd</a>, <a href="../../slides/05_Bioconductor/lab/GRL.Rmd" target="_blank">GRL.Rmd</a>

### Annotations

<a href="../../slides/05_Bioconductor/03_Annotation.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/03_Annotation.html#1')
```

<a href="../../slides/05_Bioconductor/lab/annotation.R" target="_blank">annotation.R</a> - working with annotation packages. <a href="../../slides/05_Bioconductor/lab/biomaRt.R" target="_blank">biomaRt.R</a> - interacting with biomaRt databases.


### Summarized Experiment

<a href="../../slides/05_Bioconductor/04_SummarizedExperiment.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/04_SummarizedExperiment.html#1')
```

<a href="../../slides/05_Bioconductor/lab/SummarizedExperiment.R" target="_blank">SummarizedExperiment.R</a> - Creating SummarizedExperiment. <a href="../../slides/05_Bioconductor/lab/airway_data.zip" target="_blank">airway_data.zip</a> - data for summarizedExperiment.

