---
title: "RNA-seq"
# linktitle: "1. Introduction to RStudio and Images"
date: "2021-03-03"
class_date: "2021-03-03"
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 8
type: docs
# weight: 1 
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Slides

### RNA-seq overview

<a href="../../slides/06_RNA-seq/01_RNA-seq_overview.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/01_RNA-seq_overview.pdf')
```

### RNA-seq experimental dedign

<a href="../../slides/06_RNA-seq/02_RNA-seq_expdesign.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/02_RNA-seq_expdesign.pdf')
```

### RNA-seq QC and alignment

<a href="../../slides/06_RNA-seq/03_RNA-seq_qc_alignment.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/03_RNA-seq_qc_alignment.pdf')
```

### RNA-seq quantification

<a href="../../slides/06_RNA-seq/04_RNA-seq_quantification.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/04_RNA-seq_quantification.pdf')
```

### RNA-seq normalization

<a href="../../slides/06_RNA-seq/05_RNA-seq_normalization.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/05_RNA-seq_normalization.pdf')
```

### RNA-seq batch effect

<a href="../../slides/06_RNA-seq/06_RNA-seq_batch.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/06_RNA-seq_batch.pdf')
```

### RNA-seq differential expression

<a href="../../slides/06_RNA-seq/07_RNA-seq_diffexpression.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/07_RNA-seq_diffexpression.pdf')
```

<a href="../../slides/06_RNA-seq/07a_Intro.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/07a_Intro.pdf')
```

<a href="../../slides/06_RNA-seq/07b_limma.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/06_RNA-seq/07b_limma.pdf')
```
