---
title: "Epigenomics, methylation"
# linktitle: "1. Introduction to RStudio and Images"
date: "2021-04-07"
class_date: "2021-04-07"
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 12
type: docs
# weight: 1 
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Slides

### Epigenomics intro

<a href="../../slides/10_Epigenomics/01_Epigenomics_intro.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/10_Epigenomics/01_Epigenomics_intro.pdf')
```

### Epigenomic enrichment

<a href="../../slides/10_Epigenomics/02_Epigenomics_enrichment.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/10_Epigenomics/02_Epigenomics_enrichment.pdf')
```

<a href="../../slides/10_Epigenomics/lab/Enrichment_Regions.Rmd" target="_blank">Enrichment_Regions.Rmd</a> (<a href="../../slides/10_Epigenomics/data/Horvath_cpg_all.bed" target="_blank">Data</a>),

### Hidden Markov Models, genome segmentation overview

<a href="../../slides/10_Epigenomics/03_HMM.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/10_Epigenomics/03_HMM.pdf')
```

<a href="../../slides/10_Epigenomics/lab/Segmentation_demo.R" target="_blank">Segmentation_demo.R</a>

### Array-based methylation technologies and analysis

<a href="../../slides/11_Methylation/02_Methylation_arrays.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/11_Methylation/02_Methylation_arrays.pdf')
```

<a href="../../slides/11_Methylation/03_Methylation_minfi.pdf.pdf" target="_blank">Additional: minfi methylation pipeline, PDF</a>

### Sequencing-based methylation technologies and analysis

<a href="../../slides/11_Methylation/04_Methylation_sequencing.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/11_Methylation/04_Methylation_sequencing.pdf')
```

### Cell type deconvolution

<a href="../../slides/11_Methylation/05_Methylation_deconvolution.pdf" target="_blank">Lecture slides in PDF format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/11_Methylation/05_Methylation_deconvolution.pdf')
```








