---
title: "Filtering"
# author: "Mikhail Dozmorov"
date: '`r Sys.Date()`'
output:
  pdf_document:
    toc: no
  html_document:
    theme: united
    toc: yes
---

```{r setup, echo=FALSE}
# Set up the environment
library(knitr) 
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=T, dpi=100, out.width=700)
options(replace.assign=TRUE, width=220)
options(stringsAsFactors = FALSE)
set.seed(1)
```

### genefilter & collapseRows - removing genes with certain properties before DE analysis

Filtering any unimportant genes before differential expression (DE) analysis is quite desirable because it increases the statistical power of your differential expression analysis. In other words, a smaller gene set suffers less of a hit from multiple testing correction. 

Another potential use for filtering is reducing the feature space for prediction analysis. This will cause models to be trained faster and probably to have less variance.

Of course, the tradeoff in both cases is that the genes you filter may be false negatives: that they may be truly differentially expressed or may carry unique predictive power. Filtering also reduces variability needed for proper functionality of the `limma` package [https://www.ncbi.nlm.nih.gov/pubmed/20460310](https://www.ncbi.nlm.nih.gov/pubmed/20460310).

```{r}
library(ALL)
library(genefilter)
data(ALL)
```

Looks like log2-transformed, not normalized

```{r}
summary(exprs(ALL)[,1:3])
```

`kOverA` filters out probes which do not have at least "k" experiments in which they are over expression level "A". The idea here is that  the variability of low-expressed genes is harder to distinguish from the baseline noise level.

```{r}
ka.filter <- filterfun(kOverA(5, 4.0))
ix.ka <- genefilter(exprs(ALL), ka.filter)
nrow(exprs(ALL))
sum(ix.ka)
```

you could now alter the original ALL structure by

```{r}
filtered <- exprs(ALL)[ix.ka ,]
```

`nsFilter` stands for "non-specific filter". Its main purpose is to remove probes which do not map to genes or probes which do not have GO categories.

```{r}
filtered <- nsFilter(ALL, require.entrez=T)
filtered$filter.log
filtered$eset # The new ExpressionSet
```

Lots of other filters exist. For example, you can  filter by Anova or T-tests with "Anova" or "colttests" filters. See some more options:

```{r}
ls("package:genefilter")
```

`collapseRows` from the WGCNA package serves a similar purpose. It takes an expression matrix, a vector of probe IDs, and a vector of corresponding category IDs (usually Entrez Gene or Ensembl IDs) and collapses all the probes mapping to a given gene into one row.

It has many possible methods for collapsing the rows (in fact, an entire [paper](http://www.biomedcentral.com/1471-2105/12/322) was written just about this one function. By default, rows are collapsed by max mean. 

```{r}
library(WGCNA)
library(hgu95av2.db)

M <- exprs(ALL)
probes <- rownames(M)
xx <- as.list(hgu95av2SYMBOL)
symbols <- unlist(xx[probes])

clr <- collapseRows(M, symbols, probes, method="MaxMean")
names(clr)
M.genes <- clr$datETcollapsed
dim(M)
dim(M.genes)
M[1:5, 1:5]
M.genes[1:5, 1:5]
```

